---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(width = 180)
```

<!-- badges: start -->
[![tic](https://github.com/cynkra/cynkrathis/workflows/tic/badge.svg?branch=master)](https://github.com/cynkra/cynkrathis/actions)
[![Coverage status](https://codecov.io/gh/cynkra/cynkrathis/branch/main/graph/badge.svg)](https://codecov.io/github/cynkra/cynkrathis?branch=main)
[![CRAN status](https://www.r-pkg.org/badges/version/cynkrathis)](https://www.r-pkg.org/badges/version/cynkrathis)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CodeFactor](https://www.codefactor.io/repository/github/cynkra/cynkrathis/badge)](https://www.codefactor.io/repository/github/cynkra/cynkrathis)
<!-- badges: end -->

# cynkrathis

Package to simplify project scaffolding and common {renv} related tasks.

## cynkra RSPM snapshots

cynkra makes use of certain RStudio Package Manager (RSPM) snapshots across projects.
Usually each R Version is tied to a snapshot near it's release date.
If a snapshot is considered unstable due to certain R package version clashes, additional snapshots for specific R versions can be listed.

```{r }
head(cynkrathis::get_snapshots(), 10)
```

## Initialize {renv} with a given RSPM snapshot

By default `init_renv()` will select the `"recommended"` snapshot of the R version which is currently used.

```{r, eval=FALSE}
cynkrathis::init_renv()
```

## Miscellaneous helpers

See the grouped [pkgdown reference](https://cynkrathis.cynkra.com/reference/index.html).
